### 2022-05-20 omicron-generation.tex
think i'm done

Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron variants from the Netherlands between 13 and 26 December 2021.
The study found shorter mean serial intervals (3.5 vs 4.1 days) and mean incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly Omicron during the study period) than without (mostly Delta), but did not consider dynamical biases caused by growth-rate differences in their inference:
during this period, the Omicron cases were increasing, whereas the Delta cases were decreasing.
Here, we take the epidemiological context in the Netherlands during the study period into account to provide corrected estimates for the incubation periods and generation-interval distributions of the Delta and Omicron variants.
We show that using up-to-date generation-interval distributions is critical to accurately estimating the reproduction advantage (i.e., the ratio between the reproduction numbers of the invading and resident variants) of emerging SARS-CoV-2 variants.

@@

In order to accurately estimate incubation-period and generation-interval distributions of the Delta and Omicron variants, we have to take their epidemiological dynamics---in particular, the differences in their growth rates---into account.
To estimate the growth rates differences of the Delta and Omicron variants, we first estimate the number of COVID-19 cases caused by each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week. 
We note that the proportion of Delta and Omicron variants detected is reported by the date of sampling, whereas the case data are reported by the date of reports, meaning that there is some delay between the two data.
For simplicity, we do not account for this delay, which in turn affects our growth-rate estimates; 
instead, we later perform sensitivity analysis to assess how growth rates affect the inferences of the incubation-period and generation-interval distributions.
We also do not account for uncertainties around the estimates of the proportion of each variant---almost 2000 samples were tested on each week between the week of November 28, 2021, and the week of January 23, 2022, making the uncertainty negligible.

@@

To obtain confidence intervals on the estimated time-varying growth rates, we generate 1000 parameter sets by resampling spline coefficients from a multivariate normal distribution using the estimated variance-covariance matrices.
We calculate time-varying growth rates from each parameter set and use equi-tailed quantiles to generate 95\% confidence limits. 
% \swp{Sam is calling this ``simultaneous confidence intervals'' but not sure what he means. Need to ask again.}

@@

In Supplementary Figure S2, we present generation-interval estimates that are further stratified by the week of infectors' symptom onset (13--19 December 2021 and 20--26 December 2021).
While we generally estimate shorter mean generation intervals for the Omicron variant, but the differences are unclear across all stratification, except for within-household transmission pairs during week 50 (13--19 December 2021).
We also estimate a reduction in the mean forward generation intervals from week 50 (13--19 December 2021) to week 51 (20--26 December 2021), especially for the Delta variant.

@@

We compare estimates of the forward incubation-period and generation-interval distributions of the Delta and Omicron variants from the Netherlands in late 2021 and early 2022.
The original analysis detailing the data set previously reported a shorter mean incubation period and serial interval for the Omicron variant \citep{backer2021omicron}.
Accounting for differences in epidemic growth rates, however, we find similar incubation-period distributions for both variants but a shorter (0.3--0.8 days) mean generation interval for the Omicron variant relative to that of the Delta variant.
Finally, we estimate that the transmission advantage of the Omicron variant decreased from 2.1-fold to 1.4-fold between early December and late January.
Improving generation-interval estimates by taking dynamical effects into account may improve understanding of epidemic dynamics and control measures.

@@

We primarily rely on case data to understand epidemic patterns of the Delta and Omicron variants.
In doing so, we implicitly assume that the delay between infection and reports is fixed.
However, changes in case trajectories are sensitive to testing patterns and therefore may not accurately reflect patterns of infections.
While this limitation does not affect our generation-interval estimates, our inferences of the transmission advantages of the Omicron variant should be interpreted with care.

@@

Finally, we do not account for individual-level heterogeneities, such as age, vaccination status, or previous exposure history.
In general, epidemic growth rates likely differ between infection groups (e.g., the incidence of infection caused by any variant is expected to grow faster among immunologically naive individuals), and these growth-rate differences need to be accounted for to accurately estimate epidemiological delay distributions, including the incubation-period and generation-interval distributions.
We are not able to perform stratified analyses because individiaul-level information was not publicly available.
Therefore, while we estimate unclear differences in incubation-period distributions between Delta and Omicron infections, controlling for other covariates, such as age and immune history, may help better characterize differences in Delta and Omicron infections.

>>>>

Here, we re-analyze serial-interval data collected by \citep{backer2021omicron}, representing within- and between-household transmissions of the Delta and Omicron variants from the Netherlands between 13 and 26 December 2021.
The study found shorter mean serial intervals (3.5 vs 4.1 days) and mean incubation periods (3.2 vs 4.4 days) for transmission pairs with S-gene target failure (mostly Omicron during the study period) than without (mostly Delta), but did not consider dynamical biases caused by growth-rate differences in their inference:
during this period, the Omicron cases were increasing, whereas the Delta cases were decreasing.
We take the epidemiological context in the Netherlands during the study period into account to provide corrected estimates for the incubation periods and generation-interval distributions of the Delta and Omicron variants.
We show that using up-to-date generation-interval distributions is critical to accurately estimating the reproduction advantage (i.e., the ratio between the reproduction numbers of the invading and resident variants) of emerging SARS-CoV-2 variants.

@@

In order to accurately estimate incubation-period and generation-interval distributions of the Delta and Omicron variants, we have to take their epidemiological dynamics---in particular, the differences in their growth rates---into account.
To estimate the growth rates differences of the Delta and Omicron variants, we first estimate the number of COVID-19 cases caused by each variant by multiplying reported weekly numbers of cases by the proportion of Delta and Omicron variants detected---we use weekly time series to smooth over patterns of testing and reporting within each week. 
We note that the proportion of Delta and Omicron variants detected is reported by the date of sampling, whereas the case data are reported by the date of reports, meaning that there is some delay between the two data.
For simplicity, we do not account for this delay, which in turn affects our growth-rate estimates; 
instead, we later perform sensitivity analysis to assess how growth rates affect the inferences of the incubation-period and generation-interval distributions.
We also do not account for uncertainties around the estimates of the proportion of each variant---almost 2000 samples were tested on each week between the week of November 28, 2021, and the week of January 23, 2022, making this uncertainty negligible.

@@

To obtain confidence intervals on the estimated time-varying growth rates, we generate 1000 parameter sets by resampling spline coefficients from a multivariate normal distribution using the estimated variance-covariance matrices.
We calculate time-varying growth rates from each parameter set and use equi-tailed quantiles to generate 95\% confidence limits. 
We note that this method of calculating confidence intervals gives point-wise confidence intervals, meaning that the confidence intervals give 95\% coverage for the set of estimates at each time point;
these intervals are narrower than simultaneous confidence intervals, which give 95\% coverage for the set of estimated curvatures across the time period \citep{simpson2017}.

@@

In Supplementary Figure S2, we present generation-interval estimates that are further stratified by the week of infectors' symptom onset (13--19 December 2021 and 20--26 December 2021).
While we generally estimate shorter mean generation intervals for the Omicron variant, but the differences are unclear across all stratification, except for within-household transmission pairs during week 50 (13--19 December 2021).
We also estimate a reduction in the mean forward generation intervals from week 50 (13--19 December 2021) to week 51 (20--26 December 2021), especially for the Delta variant;
this decrease in the mean generation interval is likely associated with the lockdown.

@@

We compare estimates of the forward incubation-period and generation-interval distributions of the Delta and Omicron variants from the Netherlands in late 2021 and early 2022.
The original analysis detailing the data set previously reported a shorter mean incubation period and serial interval for the Omicron variant \citep{backer2021omicron}.
Accounting for differences in epidemic growth rates, however, we find similar incubation-period distributions for both variants but a shorter (by 0.3--0.8 days) mean generation interval for the Omicron variant relative to that of the Delta variant.
Finally, we estimate that the transmission advantage of the Omicron variant decreased from 2.1-fold to 1.4-fold between early December and late January.
Improving generation-interval estimates by taking dynamical effects into account may improve understanding of epidemic dynamics and control measures.

@@

There are several limitations to our analysis.
First, we primarily rely on case data to understand epidemic patterns of the Delta and Omicron variants.
In doing so, we implicitly assume that the delay between infection and reports is fixed.
However, changes in case trajectories are sensitive to testing patterns and therefore may not accurately reflect patterns of infections.
While this limitation does not affect our generation-interval estimates, our inferences of the transmission advantages of the Omicron variant should be interpreted with care.

@@

We do not account for individual-level heterogeneities, such as age, vaccination status, or previous exposure history.
In general, epidemic growth rates likely differ between infection groups (e.g., the incidence of infection caused by any variant is expected to grow faster among immunologically naive individuals), and these growth-rate differences need to be accounted for to accurately estimate epidemiological delay distributions, including the incubation-period and generation-interval distributions.
We are not able to perform stratified analyses because individiaul-level information was not publicly available.
Therefore, while we estimate unclear differences in incubation-period distributions between Delta and Omicron infections, controlling for other covariates, such as age and immune history, may help better characterize differences in Delta and Omicron infections.

Finally, there are several sources of biases in serial interval data that we did not consider.
For example, the direction of transmission is particularly difficult to establish for SARS-CoV-2 due to pre-symptomatic transmission.
Other sources of information, such as exposure history and positive test results, can help resolve uncertainties but are imperfect.
Serial interval data also depend on the ability of infected individuals to accurately recall when their symptoms started.
Future studies may explore how these biases affect the inference of generation intervals from serial intervals.
